How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-04-19 docs/manuscript.Rmd
WIP miscellaneous edits
- adding in out sessions to some figs
- sub commands to others
- work in progress!

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib"]
# bibliography_main: ["bib/refs_main.bib"]
# bibliography_supp: ["bib/refs_supp.bib"]
---

@@

```{r fig 1, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M1_sankey.tiff"))
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution; all parameter values and sources are listed in Table 1.*

@@

## Data and code 
All analyses were done in R and using these packages (list packages with commas and citations). All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo. The raw bite patient data at the national level is maintained by IPM. This database was last queried on *date here* for these analyses. Anonymized raw data from this database are available upon request from IPM. Anonymized raw data collected from the Moramanga District are shared in the archived repository.

@@

## Main findings
Travel time to the closest ARMC was a strong predictor of reported bite incidence in Madagascar. Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP. Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP. We estimate that expanding PEP access has the potential to greatly reduce deaths due to rabies, with expansion to an additional 200 {inline} clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`. However, our model predicts that these effects saturate, and that even given maximum expansion of PEP (i.e. to all major health centers (n = 1676) in the country), between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally, due to the distribution of clinic locations across the country and the predictions of maximum reported bite incidence even given low travel times. In addition, vial demand slightly outpaces additional deaths averted, with more vials needed per averted death as clinics are added. However, after a certain number of clinics are added, rather than large increases in reported incidence and thus vial demand, we estimate shifts in the clinic locations where bites are reported to.

## Limitations
We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008]. In the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel. In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that impact an individual’s access to PEP.

Importantly,the travel time estimates from the MAP friction surface underestimate true travel times reported by patients. This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018]. In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads. Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune. Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access.

We included a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times. This random component of the model improved fit to the data, suggesting that variation at the clinic level (i.e. differences in clinic functioning such as vaccine availability) and at the catchment level (i.e. differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand. In addition, we found that the scale of data used to fit these models is likely an important factor in estimating the impact of spatial access. The commune level data from the Moramanga District showed a steeper decline in reported bite incidence with travel times, and higher estimates of bite incidence at lower travel times, resulting in lower burden estimates overall, but a greater concentration of burden in areas with high travel times. However, we lacked data from other catchments at the commune scale, so it is difficult to say how broadly these patterns apply. The district model fitted to the national data therefore likely shows the more conservative estimate of the impact of current and expanded PEP access on human rabies deaths.

Our modeling framework makes strong assumptions about baseline rabies exposure incidence, based on a previous study from the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018] and a global study of rabies burden[@whorabiesmodelingconsortiumPotentialImpactImproved2019]. As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into these estimates to generate a mean estimate of burden. Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same. While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies exposure incidence. Thus, our estimates are more representative of patterns of risk than quantitative differences in burden, and spatial estimates of rabies exposure incidence would be necessary to generate more precise subnational estimates of burden.

We make several simplifying assumptions regarding PEP provisioning and efficacy. First, we assume perfect compliance of patients and efficacy of PEP. Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018]. We also assume that patients report to their closest clinic, but in our data approximately 10% of patients reported to other clinics. As we may not be assigning reported bites to the right clinic, we may potentially be biasing our estimates of vial sharing and throughput at the clinic level.

Claire's MsC + Helene's comments here!
Other ways to designate clinics
Shortages!

The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported). Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems.

Current practices of provisioning PEP reactively to Category I exposures. Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped [@rajeevHealthcareUtilizationProvisioning2018].

Ultimately, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to care, and human behavior. We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning, and the policy implications of these exceptions.

## Box 1: case studies of exceptions to the travel time effects

### Anosibe A'nala
In the district of Anosibe A'nala (population of ~ xx people) south of the Moramanga District, travel times to the nearest ARMC (the Moramanga ARMC) are high and highly variable, often exceeding 24 hours.  Despite this, we see a high incidence of reported exposures. We interviewed xx% of these patients when they reported to the clinic, and about xx% {inline} of persons interviewed were loosely defined Category I exposures, members who were close to our touched a person that died from suspected cases. An xx% of them were suspected cases. And no cases were confirmed. The case of AAA points to three key issues: 1) High burden of rabies in this district, we're probably seeing the tip of the iceberg. 2) people when aware of the risk do travel (in many cases walking 3 days to attend the clinic) and when surveillance efforts target cat I around deaths too. Potential for awareness and outreach to go far. 3) This is also where our travel time estimates fail us. Because while there is a road, road speeds are often < 10km per hour and people hitch rides with transport trucks. This is also highly seasonal, during the rainy season it's almost impossible to travel other than by foot. And most people use multiple modes of transport depending on cost and availability (i.e. boat, on foot, by camion).

### Antananarivo (too identifiable?)
A middle aged taxi driver died of rabies in suburban Antananarivo. Bitten by an unknown dog. Promptly reported to a clinic where he was reffered to IPM. His daughter, a prospective medical student, and other family members urged him to seek the vaccine but he did not in the end seek care. He developed symptoms 2 weeks later and died within a few days. The case was confirmed by the NRL. This case points to the other side of the health-seeking behavior coin from AAA. This man had the resources, was aware of the risk, and was reffered by the hospital (and sought care immediately). But ultimately, it's difficult to predict human behavior.

## Broader context
While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings (1 - 5 subnationally, on average 1.5, see [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009]. In particular, the estimate of incidence of deaths for the Moramanga District fell within the range of incidence estimated from a contact tracing study in this district (1.35 {inline} vs. 1-3 deaths {inline} per 100,000 persons and reporting 72 vs 85 {inline}, [@rajeevHealthcareUtilizationProvisioning2018]). At the national level, they are in line with estimates of burden based on a more standard decision tree approach [@whorabiesmodelingconsortiumPotentialImpactImproved2019].

<!-- TO DO: WHAT'S THE PT I'M TRYING TO MAKE HERE? -->
As bites due to other reasons than rabies can be more common than genuine rabies exposures [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], reported bite data may be a poor indicator of rabies transmission, particularly in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population). Surveillance instead...Integrated bite case management and clinic-based surveillance approaches have been established as feasible tools that could accomplish this [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017], and ARMC could act as sentinels for rabies surveillance across the country. Ultimately, these methods rely on exposed persons to report to clinics, and as such, expanding access to PEP could strengthen surveillance.

@@

In most rabies endemic countries, access to PEP is geographically limited and/or costly to patients [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].  Currently, each clinic in Madagascar serves on average a catchment population of 750,000 persons {inline}, compared to Tanzania where catchment sizes are about half of that. Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons. Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine upon demand. However, once the more Vero cell vaccine was implemented, vaccine provisioning to all clinics became too costly, resulting in provisioning to a subset of clinics [@andriamandimbySurveillanceControlRabies2013]. However, Gavi investment can remove this cost barrier. If this results in expansion to underserved areas and strengthens existing provisioning and supply chain mechanisms, then rabies burden could be greatly reduced. Given existing challenges in data collection and monitoring of how and to whom PEP is provisioned, strong monitoring and evaluation systems will be necessary to ensure these vaccines are being used judiciously and equitably.

Ultimately even given expanded access, achieving 'the last mile', i.e. preventing deaths due to rabies in the most remote populations will require a disproportionate level of resources and may not be achievable through PEP alone [@delriovilasTribulationsLastMile2017]. IBCM programs and risked based assessments are one way to tackle the issue of managing PEP resources effectively even as demand outpaces reductions in burden [@rysavaPathRabiesElimination2018]. Combining PEP provisioning with mass dog vaccination will be key to interrrupting transmission, and eliminating deaths due to rabies. These same issues of access, however apply to both dog vaccination programs and surveillance systems, and integrating spatial heterogeneities in vaccination and detection will be critical to understand how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015].

<!-- TO DO -->
citing giorbhail paper!
What are the key pts!

## Conclusion {inline these stats}
Geographic access to PEP is a strong predictor of reported bite incidence in Madagascar and expanding PEP availability to more clinics across the country could greatly reduce human rabies deaths. Based on our modeling framework, a more equitable target may be aiming to minimize the population living within 3-5 hours travel time of a clinic, but due to broad patterns of remoteness across Madagascar, even when all primary health facilities are providing PEP, approximately 10 % {inline} of people live > 3 hours away from a clinic. Expanding PEP to clinics that target currently underserved populations may be the best strategy for preventing human rabies deaths. If expanded only to one clinic per district (N = 114), an additional 40% of deaths could be averted. Expansion of PEP can greatly reduce risk and inequities in access to this life-saving vaccine, but ultimately, interrupting transmission of rabies through mass dog vaccination programs will be the most equitable way to prevent, and eventually eliminate, rabies deaths.

>>>>

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
# bibliography_main: ["bib/refs_all.bib"]
# bibliography_supp: ["bib/refs_pkgs.bib"]
---

@@

```{r fig 1, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M1_sankey.jpeg"))
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution; all parameter values and sources are listed in Table 1.*

@@

## Data and code 
All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}. All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo. The raw bite patient data at the national level is maintained by IPM. This database was last queried on September 19, 2019 for these analyses. Anonymized raw data from this database are available from IPM following insitutional data transfer protocols. Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.

@@

Notes from Claire's MSc
22 ARMC ask for fees (consultation fees to fees for guaze etc for the form. ranging from 2000 Ar - 10000 Ar in some cases, 0.50 - 3,00 USD)
60% would vaccinate cat I  exposures
3 ARMC were out of stock 
282000 Euros a year on 23800 + 9500 vacc vials, 2000 RIG vials.

## Main findings
Travel time to the closest ARMC was a strong predictor of reported bite incidence in Madagascar. Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP. Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP. We estimate that expanding PEP access has the potential to greatly reduce deaths due to rabies, with expansion to an additional 200 {inline} clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`. However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP (i.e. to all major health centers (n = 1676) in the country), between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally, due to the distribution of clinic locations across the country. After a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to. Vial demand reflects these patterns as well, with more vials needed per death averted as clinics are added, due to the demand for PEP from patients not genuinely exposed to a rabid animal.

## Limitations
The travel time estimates from the MAP friction surface underestimate true travel times reported by patients. This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018]. In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads. Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune. Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access, or by collecting data on patient reported travel times. While patient reported travel times may not be precise, they will likely be more reflective of the range of travel realities that people experience.

The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported). Improving data submission and quality from clinics across the country will be important to better understand patterns of health seeking behavior and access, but also to monitor how well clinics are functioning and to develop better provisioning systems. In addition, we found that the scale of data used to fit these models is likely an important factor in estimating the impact of spatial access. The commune level data from the Moramanga District showed a steeper decline in reported bite incidence with travel times and higher estimates of bite incidence at lower travel times, resulting in lower burden estimates overall, but a greater concentration of burden in areas with high travel times. However, we lacked data from other catchments at the commune scale, and the commune model we developed to fit to the district scale data was not able to capture these same patterns from the national data. The district model fitted to the national data shows the more conservative estimate of the impact of current and expanded PEP access on human rabies deaths.

We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008]. In the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel. In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that impact an individual’s access to PEP. We included a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times. This random component of the model improved fit to the data, suggesting that variation at the clinic level (i.e. differences in clinic functioning such as vaccine availability) and at the catchment level (i.e. differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand. A recent knowledge-attitudes-practices (KAP) study of both community members in the Moramanga District and ARMC staff across the country found that both socioeconomic status and living in an urban area were predictive of knowledge scores [cite Leblanc 2019]. Outlier discussion?

Our decision tree framework makes strong assumptions about baseline rabies exposure incidence, based on a previous study from the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018] and a global study of rabies burden[@whorabiesmodelingconsortiumPotentialImpactImproved2019]. As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into these estimates to generate a mean estimate of burden. Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same. While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies exposure incidence. Thus, our estimates are more representative of patterns of risk than quantitative differences in burden, and spatial estimates of rabies exposure incidence would be necessary to generate more precise subnational estimates of burden.

We also make several simplifying assumptions regarding PEP compliance and efficacy. First, we assume perfect compliance of patients and efficacy of PEP. Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018]. We also assume that all ARMC reliably provision PEP, however Leblanc 2019 found that at the time of the KAP survey, three clinics were out of vaccine stock which likely contributes to burden. We do not account for differential risk for severely exposed patients (i.e. Category III exposures) that do not receive rabies Immunoglobulin (RIG). RIG is only available at IPM in Antananarivo, and while exposed persons may be referred to seek RIG, the majority of Category III exposures that seek care are unlikely to have received RIG. Unfortunately, we were not able to link patient records between peripheral ARMC and IPM to directly estimate this. Recent studies have shown that even in the absence of RIG, PEP is highly effective at preventing rabies deaths, and only adds a marginal benefit in terms of reducing burden but at very high cost. Therefore, it is both unlikely that availability of RIG contributes significantly to rabies burden in Madagascar, or that expansion of RIG provisioning would be a feasible option given the costs.

Ideally, many of these assumptions could either be validated by or replaced with improved data on canine rabies incidence, health-seeking behavior, and clinic functioning and provisioning of PEP. However, even with these data sources, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to and quality of care, and human behavior to predict rabies burden. We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning, and the policy implications of these exceptions.

## Box 1: case studies of the challenges of predicting health seeking behavior

### Anosibe A'nala
In the district of Anosibe A'nala (population of ~ xx people) south of the Moramanga District, travel times to the nearest ARMC (the Moramanga ARMC) are high and highly variable, often exceeding 24 hours and consisting of days of travel on foot. Despite these barriers, we see a high incidence of reported exposures (~~ xx per 100,000 persons). We interviewed xx% of these patients when they reported to the clinic, and about xx% {inline} of persons interviewed were loosely defined Category I exposures, members who were in close proximity or had touched a person that had died from suspected rabies. Approximately xx% of non Category exposures were classified as a probable rabies case based on clinical criteria. No cases were confirmed diagnostically however. The data from Anosibe A'nala points to three key insights: 
1) This remote community does suffer from rabies exposure and is likely experiencing a high burden of death that we are only seeing the tip of the iceberg for. 
2) When people are aware of the risk of death due to rabies, they are willing to travel a long time (in many cases walking 3 days to attend the clinic) to seek the vaccine. In addition, surveillance responses, i.e. when there is a suspected cases, there is often referral of these people from far away, can work and if these investigations are triggered on exposure rather than death then they could greatly help to prevent deaths. There is great potential for community outreach to go far. 
3) The travel times estimated from the friction surface are much lower than those observed. Because while there is a road, road speeds are often < 10km per hour and people hitch rides with transport trucks. This is also highly seasonal, during the rainy season it's almost impossible to travel other than by foot. And most people use multiple modes of transport depending on cost and availability (i.e. boat, on foot, by camion). So these metrics may miss key areas that are underserved and warrants improvement.

### Antananarivo 
A middle aged taxi driver died of rabies in suburban Antananarivo in 2016. Bitten by an unknown dog. Promptly reported to a clinic where he was reffered to IPM. His daughter, a prospective medical student, and other family members urged him to seek the vaccine but he did not in the end seek care. He developed symptoms 2 weeks later and died within a few days. The case was confirmed by the Naional Rabies Reference Laboratory at IPM. This case points to the other side of the health-seeking behavior coin from AAA. This man had the resources, was aware of the risk, and was reffered by the hospital (and sought care immediately). But ultimately, it's difficult to predict human behavior. Ultimately it will be difficult to acheive 100% coverage of this intervention even in the best of settings. Why we don't see people dying in other places is because of the combo of dog vacc + PEP vacc.

## Broader context

@@

The observed range of reported bite incidence from our data are in line with estimates of reported bite incidence across Africa, and largely is in line with previous model estimates of bite incidence based on national statistics (i.e. human development index & proportion of the population that is urban, see @@whorabiesmodelingconsortiumPotentialImpactImproved2019 Supplementary Materials for both a summary of data sources and model estimates). Ultimately, reported bite incidence is driven by many factors unrelated to rabies including the underlying size of dog populations, barriers to access to PEP (i.e. both financial costs and geographic access), and awareness of rabies risk. In our study, we do not attempt to quantify the total non-rabies related bite incidence, but use a range of values for the proportion of reported bites which we think are due to rabies. The approach we used here is likely less useful in settings where rabies exposures only make up a small fraction of patients reporting for PEP [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], i.e. in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population).

While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings from both empirical and modeling studies (1 - 7 subnationally, on average 2.7 - 3.1, see [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009, @whorabiesmodelingconsortiumPotentialImpactImproved2019]. Our estimates of rabies burden fell within the higher end of the range of previous estimates of burden for Madagascar [@rajeevHealthcareUtilizationProvisioning2018]. While we assumed a lower range of incidence of underlying rabies exposure, accounting for spatial variation in access and in reported bite incidence across the country resulted in predictions of increased burden, mirrored by a decrease in deaths averted under baseline provisioning. However, our results covered wide 95% prediction intervals, and improved clinic level data, as well as surveillance data on rabies exposures will be necessary to further refine these absolute estimates.

One approach to obtaining this data is integrated bite case management and clinic-based surveillance, methods that relies on follow-up of bite patients reporting to determine rabies exposure status, recommend treatment, and identify other potential persons exposed [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017; @rajeevHealthcareUtilizationProvisioning2018]. If these methods were implemented, ARMC could act as sentinels for rabies surveillance across the country, with bite patients reporting to clinics trigger investigations of suspect rabies cases. Ultimately, these methods rely on exposed persons to report to clinics, and as such, expanding access to PEP could strengthen clinic based surveillance efforts.

Our estimates of vial demand depend on countrywide adoption of the new abridged intradermal regimen recommended by WHO [@tarantolaIntradermalRabiesPostexposure2018]. While this new recommendation has been adopted by the Ministry of Health, it will be important to train clinicians and staff, as in 2019 only 15% of staff interviewed were aware of this regimen [Leblanc 2019]. In addition, the majority of staff were also not aware of the WHO classifications of exposure categories, and vaccination of Category I exposures, i.e. those not recommended PEP, is a common practice. In the Moramanga district, up to 20% of vial demand was due to these exposures [@rajeevHealthcareUtilizationProvisioning2018]. More judicious PEP provisioning could reduce vial demand. Current practices of provisioning PEP reactively to Category I exposures. Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped.

We predict that as clinics are expanded, overall throughput (i.e. the number of patients reporting daily to a clinic) will decrease. This may make provisioning PEP more challenging, as vial demand may become less predictable annually, and result in stockouts or vaccine wastage. Establishing decentralized provisioning mechanisms, for example through combining provisioning of PEP with routine childhood vaccines, or novel methods for vaccine delivery (i.e. drone delivery of emergency medicine, [@boeras2019use]) may be potential ways to mitigate these challenges. Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine from IPM upon demand. However, once the more expensive Vero cell vaccine was implemented, vaccine provisioning to all clinics became too costly, resulting in provisioning to a subset of clinics [@andriamandimbySurveillanceControlRabies2013]. This suggests that wider spread vaccine provisioning is feasible if cost barriers are removed.
